\section{Introduction}

Over the last few decades, the speed of computers has increased by orders of
magnitude, but the speed of programmers has not. It is often far more
important to quickly produce correct and robust code than to optimise code for
performance, and as computers continue to become more powerful, while humans
will essentially remain the same, this is ultimately going to become the norm.
We argue that prototyping new heuristics and algorithms for combinatorial
optimisation is one area where speed of development of correct code is already
more important than absolute performance.

In this paper we undertake a small case study as a preliminary investigation
into whether a declarative approach, specifically functional programming, is
feasible for this domain, and if it can help speeding up prototyping. Our
basic observation is that algorithms and heuristics for combinatorial
optimisation at their core have clear mathematical specifications, but that
implementation often is hampered by the need to spell out a plethora of
operational details. This is time-consuming, error prone, and ultimately
obscure the essence of the code. Thus, if combinatorial optimisation
algorithms could be prototyped by, for the most part, transliterating the
mathematical specifications, and if the resulting performance were adequate
for evaluation purposes, much would be gained already.
% Additionally, declaratively formulated code, as straightforward equational
% reasoning is valid, facilitates using techniques such as property-based
% testing \cite{quickcheck}, which has proved highly effective in many domains,
% and even formal reasoning and proving correctness.
Additionally, because the elementary, ``school-book'' reasoning principle of
substituting equals for equals is valid in declaratively formulated code
(unlike in a setting with side effects), applying property-based testing
\cite{quickcheck} (where test cases are derived automatically from stated
correctness properties), and formally proving correctness, is facilitated.

For our case study, we have opted to look at a few standard dynamic
programming algorithms, including bounded and unbounded knapsack. These have
many uses in scheduling and timetabling, and, for our purposes, are
representative of a larger class of algorithms in the domain of combinatorial
optimisation. For the declarative implementation framework, we have opted to
use the pure, lazy, functional language Haskell \cite{Haskell98Book}. Using a
pure language increases the contrast to the imperative languages commonly used
to implement this class of algorithms, making for a more interesting
comparison, while also allowing the specific advantages of working
declaratively
% Could say: "as discussed above", but being less specific means this could
% also be taken to refer to other advantages such as being able to exploit 
% laziness.
to be fully realised. Further, Haskell is supported by mature,
industrial-strength implementations, resulting in a fairer performance
comparisons.

We would like to emphasise that our aim is not to advocate any particular
functional language for prototyping combinatorial optimisation algorithms.
Rather, we are interested in exploring what advantages functional notation
supported by mature implementations can bring today. However, it is
worth noting that if these
advantages are judged to be compelling enough, functional
language implementations can, with relative ease, be leveraged for
implementing domain-specific languages that allow domain-experts interested in
working declaratively to reap the benefits of the approach without having to
become seasoned functional programmers themselves \cite{Hudak1998}.

We carry out the study by implementing each of the chosen algorithms in
Haskell, Java, and C. The implementations are then compared along a number
of dimensions, including conciseness, modularity, debugging, reasoning, and
performance. To make the comparisons meaningful, we retain the structure the
of the implementations across languages, except where we take advantage of
specific language features (such as pointers, objects, or laziness). The
implementations are further idiomatic and representative of what a
``typical'' programmer might write, without non-portable micro-optimisations.
In particular, standard libraries are used throughout for data structures and
mathematical computations, with as little as possible implemented from
scratch.

% Our findings so far suggest that functional languages supported by mature
% implementations indeed can speed up development by allowing implementations
% to stay close to specifications, taking advantage of specific language
% features such as laziness, and eliminating certain classes of errors,
% without incurring a performance penalty that is unacceptable for prototypes.
